Investigating soundscapes perception through acoustic scenes simulation.

### 2018-01-17 lafaySimulatedScenePerception.tex
mathias comments

\documentclass[12pt]{elsarticle}
\usepackage{url}
\usepackage[utf8]{inputenc}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{caption}
\usepackage{sectsty}
\usepackage{amsmath}
\usepackage{color}

@@

@@

One of the main goal of soundscape studies is to identify which components of the soundscape influence human perception \cite{aletta2016soundscape}. For example, in order to study the link between an environment (urban sound scene) and the induced human sensation (calmness), one mean is to objectify the mental representation of a specific sound scene (a calm urban sound scene). Several means have been considered in the literature to perform this task.

@@

First, a subject can be asked to assess a given sound following a given perceptual scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}. The amount of information that can be gathered strongly depends on the very nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}. That being said, those recorded stimuli  can be analyzed, what allows the experimenter to gather a coarse
grain physical description of them.

@@

We propose in this paper to consider the use of a soundscape simulator that the subject can manage to objectify his/her representation of a given sound environment. We believe that the use of such a device allows us to gain the benefits of the two above cited approaches. As the subject is asked to produces audio data (the signal of the simulated scene), it allows the experimenter to study a precise modal version of the mental representation of the subject that is both characterized semantically and physically, respectively in terms of the sound sources and their levels.

@@

We believe that the availability of a fine grain description of the sound stimuli is of great interest, because recent research studies demonstrate that the sound sources do not contribute equally to the perception of the sound scene \cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,
szeremeta2009analysis}. Thus, much attention is given to the specific contributions of the different sources on the notion of emotional quality of the scene \cite{gozalo2015relationship,ricciardi2015sound}.

@@

For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed in this article can lead to interesting outcomes as it allows the experimenter to separately study  the influence of the sound sources that compose a sound environment of interest. With such a material, not only the type of sound sources is indeed available for study, but also the exact level and audio waveform for each source together with the structural characteristics of the scene, that is the temporal distribution of the sound events.

@@

To demonstrate the potential of the proposed approach in its ability to question the human perception of the sound environment, we study in this paper the notion of perceptual pleasantness of urban environmental scenes. Results and outcomes of a series of three experiments that foster on the use of the simulator are studied in order to better comprehend how different sound sources typically present in a urban scene impact pleasantness:

@@

\begin{enumerate}
\item \emph{experiment 1.a, simulation}:  the subjects use the soundscape simulator to produce ideal / non ideal soundscapes that are considered as material for the following experiments
\item \emph{experiment 1.b, pleasantness evaluation}: the subjects judge the pleasantness of the simulated scenes on a semantic scale
\item \emph{experiment 2, pleasantness evaluation after modification of the scenes}: the subjects judge the pleasantness of the simulated scenes on a semantic scale as in 1.b, but some scenes are modified beforehand, \ie~some specific sounds classes that are identified as having a significant impact on perceived pleasantness are removed.
\end{enumerate}

@@

In our approach, the simulator developed for this study\footnote{simScene, available online \url{https://bitbucket.org/mlagrange/simscene}} gives a monophonic representation of the scene, with the added benefit of a wider range of available sound sources and scheduling parameters in order to provide outputs which are both expressive and useful for analysis.

@@

The remaining of the paper is organized as follows: the soundscape simulator \emph{simScene} is introduced in Section \ref{sec:simulator}. The experiments  1.a \emph{simulation} and  1.b \emph{pleasantness evaluation} are presented in Section \ref{sec:simulation}. Experiment 2 \emph{pleasantness evaluation after modification of the scenes} is presented in Section \ref{sec:modification}. Conclusions and discussion about future work follow in Section \ref{sec:conclusion}.

@@

\emph{Simscene} is an online digital audio tool whose first version has been developed as part of the HOULE project\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}. It has been designed to run on the popular web browsers \emph{Chrome} and \emph{Firefox}. It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js} : \url{angularjs.org}} and the \emph{web-audio} standard that allows the manipulation of digital audio data within the browser\footnote{\emph{web-audio}: \url{www.w3.org/TR/webaudio}}. The interface for selecting the sound sources (\cf~Section~\ref{sec:simscene_ssf}) uses the popular \emph{D3.js} \cite{d32011} visualization library.

@@

\emph{Simscene} is designed as a simplified audio sequencer, with sequencing parameters specifically chosen for the generation of realistic soundscapes. To do so, the user first selects a sound source using a non verbal selection interface presented in Section~\ref{sec:simscene_ssf}. A track is then created for this sound source within the simulator interface. The user can then manipulate some parameters detailed in Section~\ref{sec:simscene_parametre} to control the time and magnitude distribution of the occurrences of the sound sources. Text fields are also available for the user to 1) name each track, 2) name the entire scene, 3) to provide some comments about the simulated scene.

@@

We choose not to include any musical content in the sound database, as the study of the pleasantness of a given style or genre of music is beyond the scope of the study.

@@

The literature usually distinguishes between:

\begin{itemize}
\item {a sound event}: an isolated sound, sparsely distributed whose acoustical characteristics may change with respect to time
\item {a sound texture}: an isolated sound of long duration whose acoustical characteristics are stable with respect to time \cite{saint1995classification}.
\end{itemize}

@@

We do so because we assume that the morphological differences between those two classes have some important consequences on the perception of the scenes.

@@

According to the Auditory Scene Analysis (ASA) theory, the brain processes separately events emerging from the different sources of a sound environment \cite{bregman1994auditory}. Several auditory streams are created, one for each sequence of events emitted by the same source \cite{carlyon2004brain}. In this case, the source is a perceptual entity as events that are not isolated are grouped into a stream.

@@

Maffiolo \cite{maffiolo_caracterisation_1999} distinguishes two distinct categorization processes, that differentiate themselves with the capacity of the listener to identify sound events. She shows that those processes leads to two abstract cognitive category respectively termed "the event sequences" and "the amorphous sequences".

@@

The event sequences are composed of salient events that can easily be recognized such as \emph{car start} or \emph{male speech}. They arise from a descriptive analysis based on the identification of the sound sources. On the contrary, the amorphous sequences are sound environments where distinct events can hardly be found, such as \emph{traffic hubhub}. They result from a holistic analysis based on global acoustic features.

@@

Concerning sound textures, \ie~sounds that have stable characteristics over time, \cite{mcdermott2011sound,mcdermott2013summary} demonstrates that the human brain can decide not to abstract the physical information available using some of the summary statistics discarding lower level information \cite{nelken2013ear}.

@@

At the light of those results, it appears that the processing of auditory information comprises some sort of decision concerning the nature of the stimuli \cite{nelken2013ear,mcdermott2013summary}. We thus consider in this study the soundscape as a "skeleton of events on a bed of textures" as coined in \cite{nelken2013ear}.

@@

We term "sample", a recording of an isolated sound, be it an event or a texture. Each sound class is implemented as a collection of samples perceived equivalently.

@@

The sound classes are organized hierarchically (\cf~Figure~\ref{fig:orgDb}) according to a structure that is close to the one of the vertical axis of the categorical organization proposed by E. Rosch \cite{rosch1978cognition}. The lower the level of abstraction, the more precise the description of the class and the more perceptually similar the sound sources. For classes with high level of abstraction that have sub classes, their collection of samples is the union of the collections of the sub classes.

@@

Accounting for the perceptual matters detailed previously, two taxonomies are built, one for event sounds and one for texture sounds. Four levels of abstraction are considered from the more generic one (level 0), to the more specific classes (level 2), leading to a taxonomy close to the one used in \cite{Salamon14}. Only three levels of abstraction are considered for the texture sounds.

@@

Original sounds have been recorded using a shotgun microphone \emph{AT8035}\footnote{\emph{Micro canon AT8035}: \url{eu.audio-technica.com}} plugged to a \emph{ZOOM H4n}\footnote{\emph{ZOOM H4n} : \url{www.zoom.co.jp/english/products/h4n}} recorder. The use of such a microphone allowed us to isolate as much as possible sound events of interest from the urban background. It also allowed us to avoid dominant sound sources while recording texture sounds by targeting distant areas with no dominant sound sources.

@@

By a voluntary design choice, the simulation tool does not allow the user to interact and control directly a specific sample. Interaction is done at the track level, a track being a sequence of samples. Several parameters are available to the subject to control the track:

@@

\begin{itemize}
\item \emph{sound level} ($dB$): for each sample, the sound levels are randomly sampled according to a normal distribution parametrized by the subject in terms of mean value and variance;
\item \emph{inter-onset spacing} (second): for the event's track only, and for each sound event sample, the inter-onset spacings are randomly sampled according to a normal distribution parametrized by the subject in terms of mean value and variance;
\item \emph{start and end time} (second): the subject set the first and last occurrence of a sample of the selected class.
\end{itemize}

@@

\begin{itemize}
\item \emph{event's fades} (seconds): for the event's track only, the subject can set a fade in / fade out duration applied to each sample;
\item \emph{global fades} (seconds): the subject can set global fade in / fade out duration applied to the entire track.
\end{itemize}

@@

The samples of texture are sequenced without time spacing, therefore the parameters \emph{event's fade} and \emph{inter-onset spacing} are not available for this kind of track.

@@

\begin{figure}[!tp]
		\myfloatalign
        \subfloat[]
        {\includegraphics[width=.8\linewidth]{gfx/SSF}\label{fig:ssf}} \par
        \subfloat[]
        {\includegraphics[width=\linewidth]{gfx/simscene}\label{fig:simscene}}
       \caption{Graphical interfaces for the selection of sound classes (a) and their sequencing (b) of the \emph{simScene} tool.}
\end{figure}

@@

Once the typology and the set of sounds are available, an important design issue is the need of a suitable way to display the sound dataset to the user. Most browsing tools are based on keyword indexing. However, for sensory experiment that study the objectivation of the mental representation of the subject, this mean is problematic as the availability of a verbal description of the sound can influence the choice of the subject and potentially induce biases in the analysis conclusions. Indeed, the subject may select \textit{a priori} sounds referenced as belonging to \emph{park} to build a \emph{calm} soundscape.

@@

Therefore, the selection interface considered in this study is text free and composed of spatially organized circles on a plane, each representing one leaf class.

@@

The Figure~\ref{fig:ssf} shows the interface used for the selection of events. The classes of the higher level of abstraction are plotted with circles. The spatial location of the circles is chosen according to the hierarchical organization of the sound database: sub-classes belonging to the same class are close to each others, and so on until the user reaches the lower level classes displayed in grey which are directly linked to a collection of samples.

@@

Each of those classes have a representative sound chosen arbitrarily by the authors in order to provide the same sound each time the user clicks on the circle. The subject can browse the database by listening to those prototype sounds. The efficiency of this interface compared to several others designs have been evaluated and the outcomes are discussed in \cite{lafay2016JAES}.

@@

As shown on Figure~\ref{fig:simscene}, the simulation interface displays a schematic of the scene under creation. Each track is represented as an horizontal band with a temporal axis. Each sample of this track is displayed by a rectangle whose height is proportional to the amplitude of the sample. For event's tracks, the horizontal spacing between those rectangles is a function of the time delay between their onsets. For tracks of texture, a unique rectangle is displayed as this kind of sounds does not allow spacing with silence. As actual amplitude and spacing values are drawn from random variables, each time the subject changes the value of a parameter, the location and height of the rectangles are updated to reflect the changes in the sequencing of the samples. The subject can listen to the resulting sound scene at any time.

@@

\begin{figure}[t]
        \includegraphics[width=\linewidth]{gfx/5.pdf}
        \caption{Experimental protocol of the simulation experiment (1.a) and the pleasantness evaluation experiment (1.b).\gl{TODO : translate}}\label{fig:xp1_2}
\end{figure}

@@

\begin{itemize}
\item \emph{experiment 1.a (simulation)}: during this experiment, subjects are asked to simulate urban sound environments, using the simulation tool \emph{Simscene}, see Section~\ref{sec:simulator}. Each of them has to create two sound environments: one ideal / pleasant and the other one, non ideal / unpleasant.
\item \emph{experiment 2.a (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes is available: respectively ideal (i) or non ideal (ni). Furthermore, this information is given by the creator of the scene. The second experimental step aims at investigating more deeply and more broadly our knowledge on the pleasantness of the simulated scenes. For that, a second group of subjects is asked to evaluate the pleasantness of each scene, on a semantic scale. This experiment has two goals:
\begin{enumerate}
\item to evaluate more precisely the respective influence of the various sources composing the scenes on the pleasantness  (i or ni) thanks to a finer quantification of the pleasantness of the scene.
\item to detect the presence of \emph{outliers} or ambiguous scenes. Indeed, throughout our analyses, the predefined hedonic properties of the scenes (i and ni) are used as reference. We thus need to ensure beforehand that no ambiguity exists between extreme cases of i and ni scenes, \ie~that the less pleasant i-scene remains above the more pleasant ni-scene.
\end{enumerate}
\end{itemize}

@@

The subjects have to simulate two urban sound environments of one minute each. They follow the instructions below:

@@

All the subjects start by designing the ideal environment; they read the second set of instructions at the end of the first experiment. subjects are completely free of their choices concerning sounds and synthesis parameters. The created sound environments must nevertheless fulfill the two following constraints :

@@

\begin{itemize}
\item the listening point of view is the one of a fixed listener;
\item the soundscape must be realistic, \ie~physically plausible. For instance, subjects are free to insert ten dogs in the soundscape but they cannot insert one dog barking every 10 milliseconds.
\end{itemize}

@@

Each simulation process is decomposed in several steps:

\begin{enumerate}
\item the simulation: the user is asked to:
\begin{itemize}
  \item  select the class of sounds
	\item  give a name to the selected class of sounds
	\item  set the parameters of the tracks related to the selected class of sounds, see Section~\ref{sec:simscene_parametre}.
\end{itemize}
\item the feedback : writing of a free form comment about the composed soundscape
\end{enumerate}

@@

\begin{itemize}
\item  point out the sound sources that were missing for the composition
\item  give a comment about the ergonomics of the simulation environment
\item   give a comment about the ergonomics of the selection tool
\end{itemize}

@@

All the subjects performed the experiment at the same time. They are equally distributed in three identical quiet rooms. They are not allowed to talk to each other during the experiment.

@@

Three experimenters (one in each room) are available during the whole duration of the experiment in order to assist subjects with potential hardware and software issues and to answer questions.

@@

44 students (14 females; averaging $21.6$ years of age, s.d. of $2.0$ years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part to the experiment. All the subjects live in Nantes, France, for at least two years.

@@

Due to temporal constraints, subjects only assess 30 seconds of the initial 1-minute simulated scenes generated in experiment 1.a (from sec. 15 to sec. 45).

@@

10 students (2 females; averaging $23.1$ years of age, s.d. of $1.8$ years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part to the experiment. All the subjects live in Nantes, France, for at least two years. None of them took part to the previous simulation experiment (experiment 1.a).

@@

A set of features is attached to each sound scene. The analysis is done on the basis of these features. A summary of these features (and the corresponding acronyms) is presented in Table ~\ref{tab:acronyme}. In order to be consistent with the evaluation of experiment 1.b, features are not computed in the whole duration of the sequences but only on their 30-second reduced version used as stimuli for experiment 1.b, see Section~\ref{sec:xp1b_plan}.

@@

\item \emph{semantic features}: we use a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene, \textit{i.e.} the sound classes that are present / absent from the scene. Each boolean $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise. The vector dimension ($n$) depends on the level of abstraction that is considered for the analysis. For instance, for the abstraction level 1 that includes 44 classes of sounds, $n = 44$.

\item \emph{sound levels}: to figure out the sound levels, we draw inspiration from the $L_{Aeq}$. In our case, it is a scalar computed from the  signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt (full scale). The level is obtained by computing the quadratic mean of the signal every second and averaging the results over the total duration of the scene. An A-filtering module processes the data before the quadratic means are computed. We note $L$, $L(E)$ and $L(T)$ the computed levels by respectively considering the whole set of samples, only the set of event samples, and only the set of texture samples.
\end{itemize}

@@

In order to evaluate the specific impact of the various sound sources on the perceived pleasantness, we put the data to the five following significance tests:

@@

\item \emph{analysis of the perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the  i- and ni-scenes during the experiment 1.a. To do so, we consider if there exist significant differences for the i- and ni-scenes of the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$. The significance is evaluated by a two-sample Student test for $\mathcal{A}_{scene}$ and by a paired-sample Student test for $\mathcal{A}_{subject}$;

@@

\item \emph{analysis of the sound levels}: the goal is to evaluate whether the sound levels differ between the i- and ni-scenes. The significance is measured with a two-sample Student test;

@@

\item \emph{influence of the sound levels on the perceived pleasantness}: the goal is to evaluate if the sound levels impact the perceived pleasantness. To do so, we consider linear correlations between these features and $\mathcal{A}_{scene}$. The Pearson correlation coefficient is used for that purpose;

@@

\subsubsection*{Analysis of the perceived pleasantness}

>>>>

\documentclass[12pt]{elsarticle}
\usepackage{url}
\usepackage[utf8]{inputenc}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage[position=top]{subfig}
\usepackage{caption}
\usepackage{sectsty}
\usepackage{amsmath}
\usepackage{color}

@@

\makeatletter
\newcommand*{\centerfloat}{%
  \parindent \z@
  \leftskip \z@ \@plus 1fil \@minus \textwidth
  \rightskip\leftskip
  \parfillskip \z@skip}
\makeatother

@@

One of the main goals of soundscape studies is to identify which components of the soundscape influence human perception \cite{aletta2016soundscape}. Establishing a link between an environment (urban sound scene) and the induced human sensation (calmness) is equivalent to instantiating the mental representation of a specific sound scene (a calm urban sound scene). Several means have been considered in the literature to perform this task.

@@

First, a subject can be asked to assess a given sound following a given perceptual scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}. The amount of information that can be gathered strongly depends on the nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}. Working with actual stimuli makes it possible to analyze them, thus allowing the experimenter to gather a coarse-grained physical description of them.

@@

We propose in this paper to consider the use of a soundscape simulator that the subject can employ to objectify his/her representation of a given sound environment. We believe that the use of such a device allows us to gain the benefits of the two abovementioned approaches. As the subject is asked to produce audio data (the signal of the simulated scene), it allows the experimenter to study a precise modal version of the subject's mental representation that is characterized both semantically (the nature of sound sources) and physically (\eg~the levels of sound sources).

@@

We believe that the availability of a fine grain description of the sound stimuli is of great interest, because recent studies demonstrate that the sound sources do not contribute equally to the perception of the sound scene \cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,szeremeta2009analysis}. Thus, much attention is given to the specific contributions of the different sources on the notion of emotional quality of the scene \cite{gozalo2015relationship,ricciardi2015sound}.

@@

For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed here can lead to interesting outcomes as it allows the experimenter to separately study  the influence of the sound sources that compose a sound environment. With such material, not only is the type of sound sources available for study, but also the exact level and audio waveform for each source together with the structural properties of the scene, that is, the temporal distribution of the sound events.

@@

To demonstrate the potential of the proposed approach in its ability to question the human perception of sound environments, we study in this paper the notion of perceptual pleasantness of urban environmental scenes. Results and outcomes of a series of three experiments that build upon the use of the simulator are studied in order to better comprehend how different sound sources typically present in a urban scene impact pleasantness:

@@

\begin{enumerate}
\item \emph{experiment 1.a, simulation}:  the subjects use the soundscape simulator to produce ideal / non ideal soundscapes that are considered as material for the following experiments;
\item \emph{experiment 1.b, pleasantness evaluation}: the subjects judge the pleasantness of the simulated scenes on a semantic scale;
\item \emph{experiment 2, pleasantness evaluation after modification of the scenes}: the subjects judge the pleasantness of the simulated scenes on a semantic scale as in 1.b, but some scenes are modified beforehand, \ie~some specific sounds classes that are identified as having a significant impact on perceived pleasantness are removed.
\end{enumerate}

@@

In our approach, the simulator developed for this study\footnote{simScene, available online \url{https://bitbucket.org/mlagrange/simscene}} only yields a monophonic representation of the scene, but that simplification comes with the benefit of a wider range of available sound sources and scheduling parameters in order to provide outputs which are both expressive and useful for analysis.

@@

The remaining of the paper is organized as follows: the soundscape simulator \emph{simScene} is introduced in Section \ref{sec:simulator}. Experiments~1.a \emph{simulation} and  1.b \emph{pleasantness evaluation} are presented in Section~\ref{sec:simulation}. Experiment~2 \emph{pleasantness evaluation after modification of the scenes} is presented in Section~\ref{sec:modification}. Conclusions and discussion about future work follow in Section \ref{sec:conclusion}.

@@

\emph{Simscene} is an online digital audio tool whose first version has been developed as part of the HOULE project\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}. It has been designed to run on the popular web browsers \emph{Chrome} and \emph{Firefox}. It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js} : \url{angularjs.org}} and the \emph{Web Audio} standard that allows the manipulation of digital audio data within the browser\footnote{\emph{Web Audio}: \url{www.w3.org/TR/webaudio}}. The interface for selecting sound sources (\cf~Section~\ref{sec:simscene_ssf}) uses the popular \emph{D3.js} \cite{d32011} visualization library.

@@

\emph{Simscene} is designed as a simplified audio sequencer, with sequencing parameters specifically chosen for the generation of realistic soundscapes. To do so, the user first selects a sound source using a non verbal selection interface presented in Section~\ref{sec:simscene_ssf}. A track is then created for this sound source within the simulator interface. The user can then manipulate some parameters detailed in Section~\ref{sec:simscene_parametre} to control the time and magnitude distribution of the occurrences of the sound sources. Text fields are also available for the user to 1) name each track, 2) name the entire scene, 3) provide free comments about the simulated scene.

@@

We choose not to include any musical content in the sound database, as the study of the pleasantness of a given style or genre of music is beyond the scope of this study.

@@

 A commonly accepted distinction consists in separating:

\begin{itemize}
\item {sound events}: isolated sounds, precisely located in time, whose acoustical characteristics may change with respect to time;
\item {sound textures}: isolated sounds of long duration whose acoustical characteristics are stable with respect to time \cite{saint1995classification}.
\end{itemize}

@@

In order to account for the possiblity that the morphological differences between those two classes may have some important consequences on the perception of the scenes, the simulation tool \emph{Simscene} follows this distinction and considers two distinct sound databases: one with classes of events only and the other with classes of textures only. These two types of sound classes also have specific scheduling procedures during the simulation process.

@@

According to the Auditory Scene Analysis (ASA) theory, the brain processes separately events emerging from the different sources of a sound environment \cite{bregman1994auditory}. Several auditory streams are created, one for each sequence of events emitted by the same source \cite{carlyon2004brain}, while events that cannot be isolated are grouped into a same stream.

@@

Maffiolo \cite{maffiolo_caracterisation_1999} distinguishes two separated categorization processes, either of which is triggered depending on the listener's ability to identify sound events. She shows that those processes lead to two abstract cognitive categories respectively termed "event sequences" and "amorphous sequences".

@@

Event sequences are composed of salient events that can easily be recognized, such as \emph{car start} or \emph{male speech}. They arise from a descriptive analysis based on the identification of the sound sources. On the contrary, amorphous sequences are sound environments where distinct events cannot be readily identified, such as \emph{traffic hubhub}. They result from a holistic analysis based on global acoustic features.

@@

Concerning sound textures, \ie~sounds that have stable characteristics over time, \cite{mcdermott2011sound,mcdermott2013summary} demonstrates that the human brain can opt for an abstract, statistical representation of the perceived information, discarding precise physical properties of the sound \cite{nelken2013ear}.

@@

In the light of those results, it appears that the processing of auditory information comprises some sort of decision concerning the nature of the stimuli \cite{nelken2013ear,mcdermott2013summary}. We thus consider in this study the soundscape as a ``\,skeleton of events on a bed of textures\,'' as coined in \cite{nelken2013ear}.

@@

We call ``\,sample\,'' a recording of an isolated sound, be it an event or a texture. Each sound class is implemented as a collection of samples judged to be perceptually equivalent.

@@

The sound classes are organized hierarchically (\cf~Figure~\ref{fig:orgDb}) according to a structure similar to the vertical axis of the categorical organization proposed by E. Rosch \cite{rosch1978cognition}. The lower the level of abstraction, the more precise the description of the class and the more perceptually similar the sound sources. For classes with a high level of abstraction that have sub classes, their collection of samples is the union of the collections of the sub classes.

@@

Accounting for the previously detailed perceptual matters, two taxonomies are built, one for event sounds and one for texture sounds. Four levels of abstraction are considered from the most generic (level 0) to the most specific classes (level 2), leading to a taxonomy close to the one used in \cite{Salamon14}. Only three levels of abstraction are considered for the texture sounds.

@@

Original sounds have been recorded using a shotgun microphone \emph{AT8035}\footnote{\emph{AT8035 shotgun microphone}: \url{eu.audio-technica.com}} plugged to a \emph{ZOOM H4n}\footnote{\emph{ZOOM H4n} : \url{www.zoom.co.jp/english/products/h4n}} recorder. The use of such a microphone allowed us to isolate as much as possible sound events of interest from the urban background. It also allowed us to avoid dominant sound sources while recording texture sounds by targeting distant areas with no dominant sound sources.

@@

By a deliberate design choice, the simulation tool does not allow the user to interact with and control directly a specific sample. Interaction is done at the track level, a track being a sequence of samples. Several parameters are available to the subject to control the track:

@@

\begin{itemize}
\item \emph{sound level} ($dB$): for each sample, the sound levels are drawn randomly following a normal distribution parameterized by the subject in terms of mean value and variance;
\item \emph{inter-onset spacing} (second): for event tracks only, and for each sound event sample, the inter-onset spacings are drawn randomly following a normal distribution parameterized by the subject in terms of mean value and variance;
\item \emph{start and end time} (second): the subject sets the start and end times between which the texture or sequence of repeated events occurs.
\end{itemize}

@@

\begin{itemize}
\item \emph{event fades} (seconds): for the event tracks only, the subject can set a fade in / fade out duration applied to each sample;
\item \emph{global fades} (seconds): the subject can set global fade in and fade out durations applied to the entire track.
\end{itemize}

@@

Texture samples are sequenced without time spacing, therefore the parameters \emph{event fade} and \emph{inter-onset spacing} are not available for this kind of track.

@@

\begin{figure}[!tp]
		\myfloatalign
        \subfloat[]
        {\includegraphics[width=.8\linewidth]{gfx/SSF}\label{fig:ssf}} \par
        \subfloat[]
        {\includegraphics[width=\linewidth]{gfx/simscene}\label{fig:simscene}}
       \caption{\emph{SimScene} graphical interfaces for the selection of sound classes (a) and their sequencing (b).}
\end{figure}

@@

Once the typology and the set of sounds are available, an important design issue is the need of a suitable way to display the sound dataset to the user. Most browsing tools are based on keyword indexing; however, for sensory experiments that study the objectivation of a subject's mental representations, this means is problematic: indeed, the availability of a verbal description of the sound can influence the subject's choice, and potentially induce biases in the analysis conclusions. For example, a subject may automatically select sounds referenced as belonging to a \emph{park} environment to build a \emph{calm} soundscape, rather than focusing on their perception.

@@

Therefore, the selection interface considered in this study is text free and designed so as to force the user to rely on listening.

@@

Figure~\ref{fig:ssf} shows the interface used for the selection of events. Each circle corresponds to a sound class, the lowest level of abstraction (leaves) colored in grey. The spatial location of those circles is chosen according to the hierarchical organization of the sound database: sub-classes belonging to the same class are close to each others, and so on until the user reaches the leaf classes, which are directly linked to a collection of samples.

@@

Each of those classes has a representative sound chosen arbitrarily by the authors in order to provide the same sound each time the user clicks on the circle. The subject can browse the database by listening to those prototype sounds. The efficiency of this interface compared to several others designs have been evaluated and the outcomes are discussed in \cite{lafay2016JAES}.

@@

As shown on Figure~\ref{fig:simscene}, the simulation interface displays a schematic of the scene under creation. Each track is represented as a horizontal strip with a temporal axis. Each sample of this track is displayed as a rectangle whose height is proportional to the amplitude of the sample. For event tracks, the horizontal spacing between those rectangles is a function of the time delay between their onsets. For texture tracks, a unique rectangle is displayed as this kind of sounds does not allow spacing with silence. As the actual amplitude and spacing values are drawn from random variables, each time the subject changes the value of a parameter, the location and height of the rectangles are updated to reflect the changes in the sequencing of the samples. The subject can listen to the resulting sound scene at any time.

@@

\begin{figure}[t]
        \includegraphics[width=\linewidth]{gfx/5.pdf}
        \caption{Experimental protocol of the simulation experiment (1.a) and pleasantness evaluation experiment (1.b).\gl{TODO : translate}}\label{fig:xp1_2}
\end{figure}

@@

\begin{itemize}
\item \emph{experiment 1.a (simulation)}: during this experiment, subjects are asked to simulate urban sound environments using \emph{Simscene} (see Section~\ref{sec:simulator}). Each of them has to create two sound environments: one ideal~/ pleasant, and the other non ideal~/ unpleasant.
\item \emph{experiment 2.a (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes is available: respectively ideal (i) or non ideal (ni). Furthermore, this information is given by the creator of the scene. The second experimental step aims at investigating more deeply and more broadly our knowledge on the pleasantness of the simulated scenes. For that, a second group of subjects is asked to evaluate the pleasantness of each scene produced during (1.a), on a semantic scale. This experiment has two goals:
\begin{enumerate}
\item to evaluate more precisely the respective influence of the various sources composing the scenes on the pleasantness  (i or ni) thanks to a finer quantification of the pleasantness of the scene;
\item to detect the presence of \emph{outliers} or ambiguous scenes. Indeed, throughout our analyses, the predefined hedonic properties of the scenes (i and ni) are used as reference. We thus need to ensure beforehand that no ambiguity exists between extreme cases of i and ni scenes, \ie~that the least pleasant i-scene remains more highly rated on average than the most pleasant ni-scene.
\end{enumerate}
\end{itemize}

@@

The subjects are asked to simulate two urban sound environments of one minute each, following these instructions:

@@

All the subjects start by designing the ideal environment; they read the second set of instructions at the end of the first experiment. Subjects are completely free of their choices concerning sounds and synthesis parameters. The created sound environments must nevertheless fulfill the two following constraints :

@@

\begin{itemize}
\item the listening point of view is that of a fixed listener;
\item the soundscape must be realistic, \ie~physically plausible. For instance, subjects are free to insert ten dogs in the soundscape but they cannot insert one dog barking every 10 milliseconds.
\end{itemize}

@@

Each simulation process is decomposed into several steps:

\begin{enumerate}
\item Simulation, where the user is asked to:
  \begin{itemize}
  \item  select sound classes,
  \item  give each of them a name,
  \item  set the parameters of the tracks related to the selected sound classes of sounds, see Section~\ref{sec:simscene_parametre}.
  \end{itemize}
\item Feedback : writing of a free form comment about the composed soundscape.
\end{enumerate}

@@

\begin{itemize}
\item  point out the sound sources that were missing for the composition;
\item  give a comment about the ergonomics of the simulation environment;
\item  give a comment about the ergonomics of the selection tool.
\end{itemize}

@@

All the subjects performed the experiment at the same time. They are equally distributed in three identical quiet rooms, and are not allowed to talk to each other during the experiment.

@@

Three experimenters (one in each room) are available during the whole duration of the experiment in order to assist subjects with potential hardware and software issues, and to answer questions.

@@

44 students (30 male, 14 female; averaging $21.6$ years of age, s.d. of $2.0$ years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part in the experiment. All the subjects had been living in Nantes, France, for at least two years at the time of the experiment.

@@

Due to temporal constraints, subjects only assess 30 seconds of the initial 1-minute simulated scenes generated in experiment 1.a (from second 15 to sec. 45).

@@

10 students (8 male, 2 female; averaging $23.1$ years of age, s.d. of $1.8$ years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part to the experiment. All the subjects had been living in Nantes, France, for at least two years at the time of the experiment. None of them took part to the previous simulation experiment (experiment 1.a).

@@

A set of features, upon which the analysis is conducted, is attached to each sound scene. A summary of those features (and the corresponding acronyms) is presented in Table ~\ref{tab:acronyme}. In order to be consistent with the evaluation of experiment 1.b, features are not computed on the whole duration of the sequences but only on their 30-second reduced version used as stimuli for experiment 1.b (Section~\ref{sec:xp1b_plan}).

@@

\item \emph{semantic features}: we use a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene, \textit{i.e.} the sound classes that are present / absent from the scene. Each boolean $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise. The vector dimension ($n$) depends on the level of abstraction that is considered for the analysis. For instance, for the abstraction level 1, that includes 44 classes of sounds ($n = 44$).

\item \emph{structural features}: while \emph{SimScene} allows us to acces a variety of information about the scene structure (such as the density of events), we only focus in this first study on the sound levels. To figure those out, we draw inspiration from the $L_{Aeq}$. In our case, it is a scalar computed from the  signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt (full scale). The level is obtained by computing the quadratic mean of the signal every second and averaging the results over the total duration of the scene. An A-filtering module processes the data before the quadratic means are computed. We note $L$, $L(E)$ and $L(T)$ the computed levels by respectively considering the whole set of samples, only the set of event samples, and only the set of texture samples.
\end{itemize}

@@

In order to evaluate the specific impact of the various sound sources on the perceived pleasantness, we run the data through the five following significance tests:

@@

\item \emph{Analysis of perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the  i- and ni-scenes during the experiment 1.a. To do so, we consider if there exist significant differences for the i- and ni-scenes of the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$. The significance is evaluated by a two-sample Student test for $\mathcal{A}_{scene}$ and by a paired-sample Student test for $\mathcal{A}_{subject}$.

@@

\item \emph{Analysis of sound levels}: the goal is to evaluate whether the sound levels differ between the i- and ni-scenes. The significance is measured with a two-sample Student test.

@@

\item \emph{Influence of sound levels on perceived pleasantness}: the goal is to evaluate if the sound levels impact the perceived pleasantness. To do so, we consider linear correlations between these features and $\mathcal{A}_{scene}$. The Pearson correlation coefficient is used for that purpose.

@@

\subsubsection*{Analysis of perceived pleasantness}
